### 2023-04-20 sandbox/STAT530_ProjectSlides.qmd
Add files via upload
Rmarkdown Quarto file for the slides -- have the introduction slides completed

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
subtitle: "STAT530 Final Project"
author: "Prayag & Tyler"
date: "4/21/2023"
format: beamer
editor: visual
header-includes:
  - \usepackage{amssymb}
---

## Presentation Outline

  * Introduction to Regression Discontinuity Design
  * [ADD LATER]
  * [ADD LATER]
  * [ADD LATER]

## Regression Discontinuity Design

**Introduction**: regression discontinuity design (RDD) is a quasi-experimental pretest-posttest design that aims to determine the causal effects of interventions by assigning a cutoff or threshold above or below which an intervention is assigned.

## Regression Discontinuity Design

**Goal**: Try to estimate the causal effect of intervention/treatment, $\tau(W=w_0)$, at $W=w_0$ on $Y$ with covariate $X$.

```{r echo=FALSE, fig.show='asis', fig.width=4.50, fig.height=3.0, fig.align='center', cache=TRUE}
N <- 10000
cc <- 3.0
X <- rnorm(n=N, mean=0, sd=1.0)
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
D <- ifelse(W>cc, 1, 0)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W)
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)
YY <- a + tau_func*DD + b*X + c*X*DD + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

##
par(cex.axis=0.75, cex.main=0.75, cex.lab=0.75)
plot(W,Y, cex=0.1, col=ifelse(D==1, "steelblue", "tomato"))
abline(v=cc, lwd=1.5)
axis(1, at=cc, labels="w_0")
legend("topleft", legend=c("Obs. w/ treatment", "Obs. w/o treatment"), 
       col=c("steelblue","tomato"), pch=c(1,1,2), cex=0.60)
par(cex.axis=0.75, cex.main=0.75)
```

## Regression Discontinuity Design

**Continuity Assumption**: Assume that the conditional expectation of potential outcome at the threshold $W=w_0$ can be approximated well from one side rather than from both sides, i.e., 
$$ \mathbb{E}(Y_i^1|W_i=w_0) = \lim_{w\downarrow w_0} \mathbb{E}(Y_i^1|W_i=w),$$
$$ \mathbb{E}(Y_i^0|W_i=w_0) = \lim_{w\uparrow w_0} \mathbb{E}(Y_i^0|W_i=w).$$

```{r echo=FALSE, fig.show='asis', fig.width=4.50, fig.height=3.0, fig.align='center', cache=TRUE}
par(cex.axis=0.75, cex.main=0.75, cex.lab=0.75)
plot(W,Y, cex=0.1, col=ifelse(D==1, "steelblue", "tomato"), ylab="Y", xlab="W")
points(W,YY,col='gray', cex=0.1, pch=2)
abline(v=cc, lwd=1.5)
axis(1, at=cc, labels="w_0")
legend("topleft", legend=c("Obs. w/ treatment", "Obs. w/o treatment", "Counterfactuals"), 
       col=c("steelblue","tomato","gray"), pch=c(1,1,2), cex=0.60)
arrows(4.5,56,3.1,39,lwd=1.5)
arrows(1.5,10,2.9,19,lwd=1.5)

```

## Regression Discontinuity Design

Therefore, under the continuity of potential outcomes assumption, we can identify the ATE, $\tau(w_0)$, at the threshold $W=w_0$ as
$$
\begin{aligned}
    \text{ATE} &= \mathbb{E}(Y_i^1 - Y_i^0|W_i=w_0) \\
               &= \lim_{w\downarrow w_0} \mathbb{E}(Y_i^1|W_i=w) - \lim_{w\uparrow w_0} \mathbb{E}(Y_i^0|W_i=w)
\end{aligned}
$$
```{r echo=FALSE, fig.show='asis', fig.width=4.50, fig.height=3.0, fig.align='center', cache=TRUE}
par(cex.axis=0.75, cex.main=0.75, cex.lab=0.75)
plot(W,Y, cex=0.1, col=ifelse(D==1, "steelblue", "tomato"), ylab="Y", xlab="W")
points(W,YY,col='gray', cex=0.1, pch=2)
abline(v=cc, lwd=1.5)
axis(1, at=cc, labels="w_0")
legend("topleft", legend=c("Obs. w/ treatment", "Obs. w/o treatment", "Counterfactuals"), 
       col=c("steelblue","tomato","gray"), pch=c(1,1,2), cex=0.60)
arrows(4.5,56,3.1,39,lwd=1.5)
arrows(1.5,10,2.9,19,lwd=1.5)

```

## Regression Discontinuity Design

A **violation** of the continuity assumption: the barely-controls and the barely-treated come from very different populations, thus our RRD is unknowingly invalidated, since we will overestimate $\tau(w_0)$.

Thus, merely plotting observed $Y$ versus $W$, we have no way of knowing if or if not the continuity assumption holds at the threshold $W=w_0$.

```{r echo=FALSE, fig.show='asis', fig.width=4.50, fig.height=3.0, fig.align='center', cache=TRUE}
N <- 10000
cc <- 3.0
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X - 0.5
D <- ifelse(W>cc, 1, 0)
X <- rnorm(n=N, mean=0, sd=0.5) + D*rnorm(n=N, mean=2.5, sd=0.5)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W) 
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)
YY <- a + tau_func*DD + b*X + c*X*DD + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

##
par(cex.axis=0.75, cex.main=0.75, cex.lab=0.75)
plot(W,Y, cex=0.1, col=ifelse(D==1, "steelblue", "tomato"), ylim=c(0,max(Y)))
points(W,YY,col='gray', cex=0.1, pch=2)
abline(v=cc, lwd=1.5)
axis(1, at=cc, labels="w_0")
legend("topleft", legend=c("Obs. w/ treatment", "Obs. w/o treatment", "Counterfactuals"), 
       col=c("steelblue","tomato","gray"), pch=c(1,1,2), cex=0.60)

```

## Regression Discontinuity Design

We can test for validity of the continuity assumption using our observed data:

  * Lack of discontinuity in desnity of assignment variable $W$ at threshold $W=w_0$ (McCrary density test).
  * Lack of discontinuity in pretreatment covariates conditional density $X|W$ at the threshold $W=w_0$ then represents empirical evidence for the continuity of expected potential outcomes so long as we test all relevant pretreatment covariates, i.e., $f(X=x|W=w)$ must be continuous in $w$ at $w=w_0$ for all $x \in X$. (Regress X on W at threshold.)

## Regression Discontinuity Design

* McCrary density test for density of assignment variable $W$ for our two sets of data:

```{r echo=FALSE, fig.show='asis', fig.width=5.00, fig.height=3.0, fig.align='center', cache=TRUE}
library(ggplot2)
library(gridExtra)
library(rdd)
cc <- 3.0
X <- rnorm(n=N, mean=0, sd=1.0)
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
D <- ifelse(W>cc, 1, 0)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W)
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)
YY <- a + tau_func*DD + b*X + c*X*DD + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

par(mfrow=c(1,2))
DCdensity(W, c=cc)
abline(v=cc)

N <- 10000
cc <- 3.0
X <- rnorm(n=N, mean=0, sd=1.0)
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
D <- ifelse(W>cc, 1, 0)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W)
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)
YY <- a + tau_func*DD + b*X + c*X*DD + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

DCdensity(W, c=cc)
abline(v=cc)
```

## Regression Discontinuity Design

* Check for empirical discontinuity in covariate $X$'s density conditional on the assignment variable $W$ at the threshold using local linear regression for both data sets:

```{r echo=FALSE, fig.show='asis', fig.width=4.50, fig.height=3.0, fig.align='center', cache=TRUE}

N <- 10000
cc <- 3.0
X <- rnorm(n=N, mean=0, sd=1.0)
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
D <- ifelse(W>cc, 1, 0)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W)
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

dcc <- 0.45
dataSIM <- data.frame(W=W, X=X, D=D, Y=Y)
data_dWm <- subset(dataSIM, W<cc & W>cc-dcc)
data_dWp <- subset(dataSIM, W>cc & W<cc+dcc)

lm.out_p <- lm(X ~ W, data=data_dWp)
lm.out_m <- lm(X ~ W, data=data_dWm)
ww_p <- data.frame(W=seq(min(data_dWp$W),max(dataSIM$W),length.out=100))
ww_m <- data.frame(W=seq(min(dataSIM$W),max(data_dWm$W),length.out=100))
xx_p <- predict(lm.out_p, newdata=ww_p, interval="confidence", level = 0.95)
xx_m <- predict(lm.out_m, newdata=ww_m, interval="confidence", level = 0.95)

p1 <- ggplot() + 
  geom_jitter(data=dataSIM, aes(x=W, y=X), color= "gray",shape=19) + 
  geom_smooth(data=data_dWm, aes(W,X),color= "darkblue",fill='blue',method="lm",se=TRUE) +
  geom_smooth(data=data_dWp, aes(W,X),color = "darkred",fill="red",method="lm",se=TRUE) + 
  labs(title="", x="W", y="X") + 
  geom_vline(xintercept=cc) + 
  geom_vline(xintercept=cc-dcc,col='purple', lty=2) + 
  geom_vline(xintercept=cc+dcc,col='purple', lty=2)

#
N <- 10000
cc <- 3.0
W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
D <- ifelse(W>cc, 1, 0)
X <- rnorm(n=N, mean=0, sd=0.5) + D*rnorm(n=N, mean=2.0, sd=0.5)
DD <- ifelse(D==1,0,1)

a=10.0; tau=10.0; b=2.5; c=0.0
tau_func <- tau*(0.5+0.4*W) 
Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

dcc <- 0.45
dataSIM <- data.frame(W=W, X=X, D=D, Y=Y)
data_dWm <- subset(dataSIM, W<cc & W>cc-dcc)
data_dWp <- subset(dataSIM, W>cc & W<cc+dcc)

lm.out_p <- lm(X ~ W, data=data_dWp)
lm.out_m <- lm(X ~ W, data=data_dWm)
ww_p <- data.frame(W=seq(min(data_dWp$W),max(dataSIM$W),length.out=100))
ww_m <- data.frame(W=seq(min(dataSIM$W),max(data_dWm$W),length.out=100))
xx_p <- predict(lm.out_p, newdata=ww_p, interval="confidence", level = 0.95)
xx_m <- predict(lm.out_m, newdata=ww_m, interval="confidence", level = 0.95)

p2 <- ggplot() + 
  geom_jitter(data=dataSIM, aes(x=W, y=X), color= "gray",shape=19) + 
  geom_smooth(data=data_dWm, aes(W,X),color= "darkblue",fill='blue',method="lm",se=TRUE) +
  geom_smooth(data=data_dWp, aes(W,X),color = "darkred",fill="red",method="lm",se=TRUE) + 
  labs(title="", x="W", y="X") + 
  geom_vline(xintercept=cc) + 
  geom_vline(xintercept=cc-dcc,col='purple', lty=2) + 
  geom_vline(xintercept=cc+dcc,col='purple', lty=2)

grid.arrange(p1, p2, ncol=2)
```

## Regression Discontinuity Design

* Both data sets passed the McCrary test, data set 2 failed the test for conditional density continuity at the threshold, thus, blind RDD is not valid for data set 2, unless you incorporate the covariate in your model (but there are caveats to adding covariates, see Calonico et al., 2018).
* **Takeaway**: RDD will produce unbiased causal estimands of intervention at the threshold, $\tau(w_0)$, without taking into account the covariates but it requires our continuity assumption for ALL relevant covariates.

```{r echo=FALSE, fig.show='hide', fig.width=3.75, fig.height=3.0, fig.align='center', cache=TRUE}
library(rdrobust)
nn <- 100
vals1 <- c()
for (i in 1:nn){
  N <- 1000
  cc <- 3.0
  X <- rnorm(n=N, mean=0, sd=1.0)
  W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
  D <- ifelse(W>cc, 1, 0)
  DD <- ifelse(D==1,0,1)

  a=10.0; tau=10.0; b=2.5; c=0.0
  tau_func <- tau*(0.5+0.4*W)
  Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

  p <- rdrobust(y=Y, x=W, c=cc, p=1, fuzzy=NULL, deriv = NULL, kernel="triangular")
  vals1 <- append(vals1, p$Estimate[1])
}

Summary<-(boxplot(vals1)$stats)
colnames(Summary)<-c("Data Set 1")
rownames(Summary)<-c("Min","First Quartile","Median","Third Quartile","Maximum")

nn <- 100
vals2 <- c()
for (i in 1:nn){
  N <- 1000
  cc <- 3.0
  W <- 4*rbeta(n=N, shape1=3.5, shape2=2) + X
  D <- ifelse(W>cc, 1, 0)
  X <- rnorm(n=N, mean=0, sd=0.5) + D*rnorm(n=N, mean=2.0, sd=0.5)
  DD <- ifelse(D==1,0,1)

  a=10.0; tau=10.0; b=2.5; c=0.0
  tau_func <- tau*(0.5+0.4*W) 
  Y <- a + tau_func*D + b*X + c*X*D + 1*W**2 + rnorm(n=N, mean=0, sd=0.25)

  p <- rdrobust(y=Y, x=W, c=cc, p=1, fuzzy=NULL, deriv = NULL, kernel="triangular")
  vals2 <- append(vals2, p$Estimate[1])
}

Summary2<-(boxplot(vals2)$stats)
colnames(Summary2)<-c("Data Set 2")
rownames(Summary2)<-c("Min","First Quartile","Median","Third Quartile","Maximum")

Summary <- cbind(Summary,Summary2)

boxplot(Summary)
abline(h=17,col='blue', lwd=1.5)
axis(2, at=cc, labels="true tao", col='blue')
```

```{r echo=FALSE, fig.show='asis', fig.width=3.75, fig.height=3.0, fig.align='center', cache=TRUE}
boxplot(Summary)
abline(h=17,col='blue', lwd=1.5)
axis(2, at=cc, labels="true tao", col='blue')
legend("topleft", legend=c("True tau"), lty=1, col="blue", lwd=1.5, cex=0.6)
```

## Regression Discontinuity Design

The estimator for the casual estimated at the assignment threshold $W=w_0$ is defined as
$$ \hat{\tau}_{\text{LLR}}(Y;W,K,b) = \hat{\alpha}_1 - \hat{\alpha}_0$$
$$ (\hat{\alpha_0},\hat{\beta_0}) = \operatorname*{arg\,min}_{\alpha_0,\beta_0} \sum_{i=1}^n 1\{w_l < W_i < w_0\}\{Y_i-\alpha_0-\beta_0(W_i-w_0)\}^2K(\cdot)$$
$$(\hat{\alpha_1},\hat{\beta_1}) = \operatorname*{arg\,min}_{\alpha_1,\beta_1} \sum_{i=1}^n 1 \{w_0 < W_i < w_u\}\{Y_i-\alpha_1-\beta_1(W_i-w_0)\}^2K(\cdot)$$
where $K(\cdot)$ is the weighting or kernel function and $b$ is the choice of bandwidth that controls the weighting scheme in this local linear regression.

### 2023-04-20 sandbox/STAT530_ProjectSlides.qmd
Add files via upload

@@
   * Introduction to Regression Discontinuity Design
+  * RRD Applied to US Congressional Elections: The Setup
   * [ADD LATER]
+  * [ADD LATER]
+
+## Presentation Outline
+
+  * **Introduction to Regression Discontinuity Design**
+  * RRD Applied to US Congressional Elections: The Setup
   * [ADD LATER]
@@
 where $K(\cdot)$ is the weighting or kernel function and $b$ is the choice of bandwidth that controls the weighting scheme in this local linear regression.
+
+## Presentation Outline
+
+  * Introduction to Regression Discontinuity Design
+  * **RRD Applied to US Congressional Elections: The Setup**
+  * [ADD LATER]
+  * [ADD LATER]
+
+## RRD Applied to US Congressional Elections: The Setup
+
+Project Motivations:
+
+  * One of the most striking facts of congressional politics in the United States is the consistently high rate of electoral success of incumbents, and the electoral advantage of incumbency is one of the most studied aspects of research on elections to the U.S.
+House (Gelman and King, 1990).
+  * The overwhelming success of House incumbents draws public attention whenever concerns arise that Representatives are using the privileges and resources of office to gain an ‘‘unfair’’ advantage over potential challengers.
+  * Def'n: The **incumbent** is the current holder of an office or position, usually in relation to an election.
+
+## RRD Applied to US Congressional Elections: The Setup
+
+Data Sources: Harvard Dataverse and MIT Election Data + Science Lab
+
+  * U.S. House 1976–2020 Elections
+      * MIT Election Data and Science Lab, 2017, "U.S. House 1976–2020"
+  * Historical Congressional Legislation and District Demographics 1972-2014
+      * Foster-Molina, Ella, 2017, "Historical Congressional Legislation and District Demographics 1972-2014"
+
+## RRD Applied to US Congressional Elections: The Setup
+
+Data Sources: Harvard Dataverse and MIT Election Data + Science Lab
+
+  * U.S. House 1976–2020 Elections (Data set 1)
+      * MIT Election Data and Science Lab, 2017, "U.S. House 1976–2020"
+  * Historical Congressional Legislation and District Demographics 1972-2014 (Data set 2)
+      * Foster-Molina, Ella, 2017, "Historical Congressional Legislation and District Demographics 1972-2014"
+
+![](HarvardDataverseImage.png){width=1.5in  fig-align="center"}
+
+## RRD Applied to US Congressional Elections: The Setup
+
+Observed variables and covariates after data wrangling:
+
+| Data Set 1           | Data Set 2   |
+| -------------------- | ------------ |
+| Year                 | Year         |
+| State                | State        |
+| District             | District     |
+| **Dem. Vote Margin** | Unemployment | 
+| Candidate            | Race         |
+|                      | Education    |
+
+**Dem. Vote Margin** = (Dem. Vote Share) - (Greatest Non-Dem. Vote Share)
+
+## RRD Applied to US Congressional Elections: The Setup
+
+![2022 Congressional Districts and Partisan Lean](CongressionalDistrictsImage.png){width=3.5in  fig-align="center"}

### 2023-04-20 sandbox/STAT530_ProjectSlides.qmd
Add files via upload

@@
   * Introduction to Regression Discontinuity Design
-  * RRD Applied to US Congressional Elections: The Setup
-  * [ADD LATER]
-  * [ADD LATER]
+  * RDD Applied to US Congressional Elections: The Setup
+  * RDD Applied to US Congressional Elections: Results
+  * RDD Applied to US Congressional Elections: Conclusions
 
@@
 
-  * **Introduction to Regression Discontinuity Design**
-  * RRD Applied to US Congressional Elections: The Setup
-  * [ADD LATER]
-  * [ADD LATER]
+  * Introduction to Regression Discontinuity Design
+  * RDD Applied to US Congressional Elections: The Setup
+  * RDD Applied to US Congressional Elections: Results
+  * RDD Applied to US Congressional Elections: Conclusions
 
@@
   * Introduction to Regression Discontinuity Design
-  * **RRD Applied to US Congressional Elections: The Setup**
-  * [ADD LATER]
-  * [ADD LATER]
+  * **RDD Applied to US Congressional Elections: The Setup**
+  * RDD Applied to US Congressional Elections: Results
+  * RDD Applied to US Congressional Elections: Conclusions
 
@@
   * Def'n: The **incumbent** is the current holder of an office or position, usually in relation to an election.
-
-## RRD Applied to US Congressional Elections: The Setup
-
-Data Sources: Harvard Dataverse and MIT Election Data + Science Lab
-
-  * U.S. House 1976–2020 Elections
-      * MIT Election Data and Science Lab, 2017, "U.S. House 1976–2020"
-  * Historical Congressional Legislation and District Demographics 1972-2014
-      * Foster-Molina, Ella, 2017, "Historical Congressional Legislation and District Demographics 1972-2014"
 
@@
 ![2022 Congressional Districts and Partisan Lean](CongressionalDistrictsImage.png){width=3.5in  fig-align="center"}
+
+## Presentation Outline
+
+  * Introduction to Regression Discontinuity Design
+  * RDD Applied to US Congressional Elections: The Setup
+  * **RDD Applied to US Congressional Elections: Results**
+  * RDD Applied to US Congressional Elections: Conclusions

### 2023-04-20 R/demographics.R
that was a grind

@@ -30,33 +30,19 @@ get_demos <- function(fn = config$data$demographics$fn,
     filter_demos()
 }
 
-test_demos <- function(df) {
-  d <- lag_any_candidates(df) |>
-    dplyr::select(year, full_district, prev_dem_margin, white = prcntWhiteAll, educ = prcntBA, unemp = prcntUnemp) |>
-    tidyr::pivot_longer(cols = c(white, educ, unemp),
+test_demos <- function(df, cols = config$analysis$colors) {
+  lag_any_candidates(df) |>
+    dplyr::select(year, full_district, prev_dem_margin, white = prcntWhiteAll, 
+                  income = meanIncome, unemp = prcntUnemp) |>
+    tidyr::pivot_longer(cols = c(white, income, unemp),
                         names_to = "demo") |>
     dplyr::filter(!is.na(value)) |>
-    dplyr::mutate(winner = ifelse(prev_dem_margin > 0, "dem", "other"))
-  
-  vals <- purrr::map_dfr(
-    seq(0.01, 0.15, by = 0.01),
-    function(bw) {
-      d |>
-        dplyr::filter(prev_dem_margin >= -bw,
-                      prev_dem_margin <= bw) |>
-        dplyr::group_by(demo, winner) |>
-        dplyr::summarize(value = mean(value),
-                         district_year_obs = dplyr::n()) |>
-        dplyr::ungroup() |>
-        dplyr::mutate(bandwidth = bw)
-    }
-  )
-  
-  lines <- ggplot(vals, aes(x = bandwidth, y = value, color = winner)) +
-    geom_line() + 
-    facet_wrap(~ demo, ncol = 1, scales = "free_y")
-  
-  ggplot(d, aes(x = prev_dem_margin, y = value, color = winner)) +
+    dplyr::mutate(Winner = ifelse(prev_dem_margin > 0, "Democrat", "Other")) |>
+    ggplot(aes(x = prev_dem_margin, y = value, color = Winner)) +
     geom_point() +
-    facet_wrap(~ demo, ncol = 1, scales = "free_y")
+    facet_wrap(~ demo, ncol = 1, scales = "free_y") +
+    theme_minimal() + 
+    scale_color_manual(values = c("Democrat" = cols$dem, "Other" = cols$rep)) +
+    labs(x = "Previous Democratic margin",
+         y = "Covariate value")
 }

### 2023-04-20 R/rdd.R

@@ -1,6 +1,6 @@
 plot_density <- function(d, fn_full, dig = 3) {
   mccrary_p <- rdd::DCdensity(d$prev_dem_margin)
-  jpeg(paste0(fn_full, "_mccrary_p_", round(mccrary_p, dig), ".jpeg"), width = 350, height = 350)
+  jpeg(paste0(fn_full, "_mccrary_p_", round(mccrary_p, dig), ".jpeg"), width = 6, height = 6, units = "in", res = 300)
   rdd::DCdensity(d$prev_dem_margin)
   dev.off()
   
@@ -16,9 +16,9 @@ plot_rdd <- function(d, fn_full,
     ggplot(aes(x = prev_dem_margin, y = dem_margin, color = prev_winner, group = prev_winner)) +
     geom_point(size = 0.5, alpha = 0.25) +
     geom_smooth(size = 2, alpha = 0.75) +
-    labs(x = "Previous Democratic margin", 
-         y = "Democratic margin",
-         color = "Previous winner",
+    labs(x = "Democratic margin, t-1", 
+         y = "Democratic margin, t",
+         color = "Winner, t-1",
          caption = paste("LATE =", late)) +
     theme_minimal() +
     scale_color_manual(values = c("Democrat" = cols$dem, "Other" = cols$rep)) +
@@ -28,10 +28,6 @@ plot_rdd <- function(d, fn_full,
     geom_vline(xintercept = 0)
   
   ggsave(paste0(fn_full, "_late_", late, ".jpeg"), p)
-  
-  jpeg(paste0(fn_full, "_late_", late, "_BASE.jpeg"), width = 1000, height = 1000)
-  plot(rdd::RDestimate(dem_margin ~ prev_dem_margin, d))
-  dev.off()
   
   NULL
 }
@@ -50,3 +46,130 @@ simple_rdd <- function(df,
   
   NULL
 }
+
+covariate_rdd <- function(df) {
+  d <- lag_any_candidates(df)
+  
+  rdd::RDestimate(dem_margin ~ prev_dem_margin | prcntWhiteAll + meanIncome + prcntUnemp, d)
+}
+
+backwards_rdd <- function(df, figs = config$paths$figures, dig = 3, cols = config$analysis$colors) {
+  d <- lag_any_candidates(df)
+  
+  mccrary_p <- rdd::DCdensity(d$dem_margin)
[…]

### 2023-04-25 Qmd/paper.qmd
start paper

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
author: "Tyler Bagwell and Prayag Gordy"
format: pdf
editor: visual
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2023-04-25 sandbox/STAT530_ProjectSlides.qmd
Add files via upload

@@
   * Introduction to Regression Discontinuity Design
-  * RDD Applied to US Congressional Elections: The Setup
+  * **RDD Applied to US Congressional Elections: The Setup**
   * RDD Applied to US Congressional Elections: Results
@@
 
-**Introduction**: regression discontinuity design (RDD) is a quasi-experimental pretest-posttest design that aims to determine the causal effects of interventions by assigning a cutoff or threshold above or below which an intervention is assigned.
+**Introduction**: Regression discontinuity design (RDD) is a quasi-experimental pretest-posttest design that aims to determine the causal effects of interventions by assigning a cutoff or threshold above or below which an intervention is assigned.
 
@@
 
-**Goal**: Try to estimate the causal effect of intervention/treatment, $\tau(W=w_0)$, at $W=w_0$ on $Y$ with covariate $X$.
+**Goal**: Try to estimate the causal effect of intervention/treatment, $\tau(W=w_0)$, at $W=w_0$ on $Y$ with (unobserved/observed) covariate $X$, where individual assignment is determined by value of $W_i$ through $D_i=1\{W_i\ge w_0\}$.
 
@@
 
-We can test for validity of the continuity assumption using our observed data:
+We empirically can test for validity of the continuity assumption using our observed data:
 
@@
 
+We use the Imbens-Kalyanaraman optimal bandwidth formula to calculate the optimal bandwidths used in our local regression estimation.
+
 ## Presentation Outline
@@
 | Data Set 1           | Data Set 2    |
-| -------------------- | ------------ |
+| -------------------- | ------------- |
 | Year                 | Year          |
@@
 | District             | District      |
-| **Dem. Vote Margin** | Unemployment | 
+| **Dem.* Vote Margin** | Unemployment | 
 | Candidate            | Race          |
@@
 **Dem. Vote Margin** = (Dem. Vote Share) - (Greatest Non-Dem. Vote Share)
+
+*Dem. = Democrat (Party)
 
@@
   * RDD Applied to US Congressional Elections: Conclusions
+
+## Sources:
+
+  1. Lee, David S.; Randomized experiments from non-random selection in U.S. House elections, Journal of Econometrics 142 (2008) 675–697
+  2. de la Cuesta, Brandon and Imai, Kosuke; Misunderstandings About the Regression Discontinuity Design in the Study of Close Elections, Annu. Rev. Polit. Sci. 2016. 19:375–96
+  3. Calonico et al.; Regression Discontinuity Designs Using Covariates; The Review of Economics and Statistics, 2019, 101(3): 442–451
+  4. McCrary, Justin; Manipulation of the running variable in the regression discontinuity design: A density test; Journal of Econometrics 142 (2008) 698–714
+
+## Audience Discussion and Questions

### 2023-04-25 Qmd/paper.qmd
add abstract

@@
 title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
+subtitle: "Final Project for STAT 530"
 author: "Tyler Bagwell and Prayag Gordy"
+date: "April 24, 2023"
+abstract: "The incumbency effect in U.S.
+elections is a subject of decades of political science research.
+Why is it, scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term?
+In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts.
+Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election.
+We found a local average treatment effect of 17.3 percentage points.
+Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
 format: pdf

### 2023-04-25 paper.qmd
move paper to main folder

---
title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
subtitle: "Final Project for STAT 530"
author: "Tyler Bagwell and Prayag Gordy"
date: "April 24, 2023"
abstract: "The incumbency effect in U.S. elections is a subject of decades of political science research. Why is it, political scientists ask, that legislative and executive bodies with such low overall approval ratings have the vast majority of its members returning term after term? In this paper, we will use a regression discontinuity design to examine the effects of party incumbency in tightly contested congressional districts. Specifically, we examine the effect of the Democratic margin in an election on the Democratic margin in the following election. We found a local average treatment effect of 17.3 percentage points. Our results further support the literature claiming a large incumbency effect, though we show its strength may be decreasing."
format: pdf
toc: true
number-sections: true
execute:
  echo: false
---

## Introduction

Incumbents in the U.S. House of Representatives are reelected at astonishing rates. Since 1964, the reelection rate has never dropped below 85%, while the approval rating of the House as a whole has generally remained around 40%, dipping to 20% in recent years. Thus, many political scientists spend their careers researching the advantages of incumbency (Gelman and King, 1990).

Many experts have criticized the apparently strong effects of incumbency. Congressmen can use government funds to visit areas of their district, meet with important political players, and send mail to constituents touting their achievements, all of which indirectly benefit their reelection campaign. Challengers enjoy none of these benefits.

Previous research has found strong effects of incumbency. Candidates running for reelection routinely raise far more money. Further, they may be able to deter challengers simply by being an incumbent. Overall, Cummings found that in similar districts, incumbent congressmen are more likely to win elections than non-incumbent congressmen from the same party.

In this paper, we will examine the incumbency effect using a sharp regression discontinuity design. In section 2, we will describe RDD and explain how it applies to the study of incumbency. Section 3 will list our results, and section 4 will explore our conclusions of the study.

## Methods
### Introduction to Regression Discontinuity Design

A regression discontinuity design (RDD) is a natural quasi-experimental design for determining causal effect of a treatment that is assigned at a cutoff. There are many examples of such treatments across diverse areas of research, such as being prescribed a drug if a biomarker exceeds a certain threshold, joining an advanced class if a test score is high enough, and receiving welfare benefits for a low enough income.

Generally, RDD attempts to estimate the local causal effect of treatment, $\tau(W = w_0)$, at $W=w_0$ on outcome $Y$ with (unobserved/observed) covariate(s) $\mathbf{X}$, where individual assignment is determined by value of $W_i$ through $D_i = \mathbb{I}\{W_i \geq w_0\}$ and where $D_i=1$ indicates the $i$'th individual received treatment. The validity of RDD on estimating the local average treatment effect (LATE) depends on the assumption that the conditional expectation of potential outcome (both treatment and control outcomes) at the threshold value $W_i=w_0$ can be well approximated from one side rather than from both sides across the assignment threshold.

If this continuity assumption holds, it indicates that individuals on both sides of the assignment threshold come from very similar populations in terms of the covariates $\mathbf{X}$ and the assigning variable $W$, such that the likelihood of individuals within this close proximity to the threshold eventually receiving treatment or remaining as a control observation is essentially random. It is this random assignment of the barely-controls and the barely-treated near the threshold that guarantees an unbiased estimation of the LATE. Additionally, our covariates $\mathbf{X}$ are necessarily pretreatment covariates, thus their values should be independent of treatment assignment. Specifically, there must be no manipulation around the cutoff; that is, an individual who is inches away from treatment must remain in the control group and may not be artificially lifted into the treatment group.

In practice, it is impossible to determine if the continuity assumption holds for a particular experimental design unless the RDD practitioner has full knowledge of the conditional population densities of the covariates $\mathbf{X}$ with respect to the assigning variable $W$, i.e., $f(\mathbf{X}|W)$, for all covariates that are relevant to $Y$. In empirical RDD, the continuity assumption can be tested with the observed data by (1) testing for the lack of discontinuity in the density of the assignment variable $W$ at threshold $W=w_0$ (McCrary density test) and (2) testing for discontinuity in the first moment of the covariates $\mathbf{X}$ conditional on $W$ at $W=w_0$. This second test is usually performed by using local linear regressions in the proximity of the line $W=w_0$. In less rigorous language, the first test is meant to determine that the odds of treatment assignment of the barely-controls and the barely-treated are purely random in the proximity of the cutoff, while the second test is to determine if the pretreatment characteristics of barely-controls and the barely-treated are indeed very similar.

The unbiased LATE on the outcome $Y$ at $W=w_0$ is itself calculated as the difference in the intercepts of two local linear regressions ($Y \sim W$) performed exclusively on either side of the cutoff $W=w_0$. Under the continuity assumption, the effects of covariates will not affect the value of the LATE calculated at $W=w_0$, thus, the covariates $\mathbf{X}$ do not need to be included in the local regression formulae in order to calculate the LATE $\tau(W=w_0)$. Analytically, the estimator for the average treatment effects is given by $$ \hat{\tau}_{\text{LLR}}(Y;W,K,b) = \hat{\alpha}_1 - \hat{\alpha}_0,$$ $$ (\hat{\alpha_0},\hat{\beta_0}) = \operatorname*{arg\,min}_{\alpha_0,\beta_0} \sum_{i=1}^n 1\{w_l < W_i < w_0\}\{Y_i-\alpha_0-\beta_0(W_i-w_0)\}^2K(\cdot),$$ $$(\hat{\alpha_1},\hat{\beta_1}) = \operatorname*{arg\,min}_{\alpha_1,\beta_1} \sum_{i=1}^n 1 \{w_0 < W_i < w_u\}\{Y_i-\alpha_1-\beta_1(W_i-w_0)\}^2K(\cdot),$$ where $K(\cdot)$ is the weighting or kernel function and $b$ is the choice of bandwidth that controls the weighting scheme in this local linear regression. In this study, we use the Imbens-Kalyanaraman optimal bandwidth formula to calculate the optimal bandwidths for local regression estimation and we used a triangular kernel for $K(\cdot)$.

One key drawback of RDD is that it can only ever give us information of the LATE at $W=w_0$ only. The external validity of RDD to determine average treatment effects for values of $W$ not equal to or at least near $w_0$ should always be called into question.

### Applying RDD to Election Outcomes and Incumbency

The data used in this study came from two primary data sets housed by the Harvard Dataverse and the MIT Election Data and Science Lab. The first data set, _U.S. House 1976–2020 Elections_, is a collection of results of US Congressional House elections between 1976 and 2020 for all congressional districts. The primary variables used in this data set were _candidate_, _candidate party_, _candidate votes_, and _total votes for election_. The second data set, _Historical Congressional Legislation and District Demographics 1972-2014_, contains a subset of the pretreatment covariates per congressional district we used for this study, namely, _unemployment rate_, _race_, and _level of education_.

The primary variable of interest for this study was _Democratic candidate vote margin_ in each election. Democratic vote margin was calculated as the difference in vote share of the Democratic candidate minus the vote share of the non-Democratic party candidate in the same election with the most votes. Unsurprisingly, for the vast majority of elections used in this study, the non-Democratic party candidate was usually a Republican. The vote share of a candidate is simply the ratio of votes obtained by a candidate versus the total amount of votes cast in the entire election for all candidates. Under this definition, the _Democratic candidate vote margin_ can take on values from $-1.0$ to $+1.0$, where a $-1.0$ vote margin indicates an election blowout by the non-Democratic candidate, a $0.0$ vote margin indicates a exact tie between the Democratic and non-Democratic candidates, and a $+1.0$ vote margin indicates an election blowout by the Democratic candidate.

In order to avoid the effects of redistricting, we excluded all elections with years ending in "2," e.g., 1982, 1992, etc., since the U.S. Constitution mandates redistricting after the decennial census that occurs in every year ending in "0." We ignored elections in which no Democrats ran, and we excluded elections where the two candidates with the highest vote margins were both Democrats, which can occur in states with jungle primaries such as California.

Regression discontinuity designs may be categorized as "sharp" or "fuzzy." Sharp designs require a hard cutoff: If observations are on the "treated" side of the line, they must receive treatment. A fuzzy design, on the other hand, adds some leeway. An observation may be on the "treated" side but decline treatment; a simple example may be a student who declines an offer to join an advanced class, even though they scored well enough on the entrance examination. In this study, we will use a sharp design. If a Democratic candidate’s margin is positive, they win the election and must take their seat, barring extremely rare circumstances such as death.

## Results
### Pre-RDD Validity Diagnostics

Before applying RDD to the election data, we first report the results of the pre-RDD diagnostic tests described in the previous section to determine how the continuity assumption holds in our election context.

First, we test the continuity of the Democratic vote margin for all $t-1$ elections in our data set. The results of the McCrary density test on the Democratic vote margin at the zero margin cutoff indicates a continuous density on both sides of the cutoff value ($\alpha = 0.05$). A plot of the McCrary density test’s density estimation is provided below.

### 2023-04-25 paper.qmd
start results

@@
   echo: false
+  message: false
+  warning: false
 ---
+
+```{r message=FALSE, warning=FALSE}
+source("main.R")
+```
 
@@
 A plot of the McCrary density test’s density estimation is provided below.
+
+```{r}
+#| fig-cap: McCrary density test and estimation for all $t-1$ Democratic vote margins.
+
+lag_any_candidates(DF) |>
+  dplyr::pull(prev_dem_margin) |>
+  rdd::DCdensity()
+```
+
+Second, we assume that the only relevant pretreatment covariates of concern for congressional district-level election outcomes are rates of unemployment, racial demographics, and level of education of a district's citizens.
+Undoubtedly, this is a gross oversimplification.
+The number of variables that affect the outcomes of close elections are innumerous, including campaign spending to the election day weather and everything in between, but collecting and analyzing all these data would be out of scope for this class project.
+Therefore, for the purposes of this small study and its given time constraints, we only consider the three aforementioned covariates.
+
+With these three pretreatment covariates chosen, we estimate the discontinuities in their conditional densities conditional on the $t-1$ Democratic election margin at the zero margin value using local linear regression and find that there are no statistically significant discontinuities in either three covariates’ densities.
+This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin.

### 2023-04-25 R/demographics.R
nicer labels

@@ -32,9 +32,9 @@ get_demos <- function(fn = config$data$demographics$fn,
 
 test_demos <- function(df, cols = config$analysis$colors) {
   lag_any_candidates(df) |>
-    dplyr::select(year, full_district, prev_dem_margin, white = prcntWhiteAll, 
-                  income = meanIncome, unemp = prcntUnemp) |>
-    tidyr::pivot_longer(cols = c(white, income, unemp),
+    dplyr::select(year, full_district, prev_dem_margin, White = prcntWhiteAll, 
+                  Income = meanIncome, Unemployment = prcntUnemp) |>
+    tidyr::pivot_longer(cols = c(White, Income, Unemployment),
                         names_to = "demo") |>
     dplyr::filter(!is.na(value)) |>
     dplyr::mutate(Winner = ifelse(prev_dem_margin > 0, "Democrat", "Other")) |>

### 2023-04-25 paper.qmd
done with 3.1

@@
 This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin.
+
+```{r}
+#| fig-cap: Three pretreatment covariates versus $t-1$ Democratic vote margin.
+
+test_demos(DF)
+```
+
+With no empirical discontinuities in any of the densities under investigation discovered, we can safely assume that the continuity assumption holds in our idealized election model here and feel comfortable applying RDD to estimating effects of incumbency.

### 2023-04-25 R/rdd.R
don't save plot

@@ -7,11 +7,11 @@ plot_density <- function(d, fn_full, dig = 3) {
   NULL
 }
 
-plot_rdd <- function(d, fn_full, 
+plot_rdd <- function(d, 
                      dig = 3, cols = config$analysis$colors) {
   late <- round(rdd::RDestimate(dem_margin ~ prev_dem_margin, d)$est["LATE"], dig)
   
-  p <- d |>
+  d |>
     dplyr::mutate(prev_winner = ifelse(prev_dem_margin > 0, "Democrat", "Other")) |>
     ggplot(aes(x = prev_dem_margin, y = dem_margin, color = prev_winner, group = prev_winner)) +
     geom_point(size = 0.5, alpha = 0.25) +
@@ -26,10 +26,6 @@ plot_rdd <- function(d, fn_full,
     scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, by = 0.25)) +
     geom_hline(yintercept = 0) +
     geom_vline(xintercept = 0)
-  
-  ggsave(paste0(fn_full, "_late_", late, ".jpeg"), p)
-  
-  NULL
 }
 
 simple_rdd <- function(df,

### 2023-04-25 R/rdd.R
don't save backwards rdd

@@ -59,7 +59,7 @@ backwards_rdd <- function(df, figs = config$paths$figures, dig = 3, cols = confi
   
   late <- round(rdd::RDestimate(prev_dem_margin ~ dem_margin, d)$est["LATE"], dig)
   
-  p <- d |>
+  d |>
     dplyr::mutate(x_winner = ifelse(dem_margin > 0, "Democrat", "Other")) |>
     ggplot(aes(x = dem_margin, y = prev_dem_margin, color = x_winner, group = x_winner)) +
     geom_point(size = 0.5, alpha = 0.25) +
@@ -74,10 +74,6 @@ backwards_rdd <- function(df, figs = config$paths$figures, dig = 3, cols = confi
     scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, by = 0.25)) +
     geom_hline(yintercept = 0) +
     geom_vline(xintercept = 0)
-  
-  ggsave(paste0(figs, "backwards_rdd" , "_late_", late, ".jpeg"), p)
-  
-  NULL
 }
 
 logistic <- function(df, cols = config$analysis$colors) {

### 2023-04-25 R/tables.R
move tables.R

@@ -1,0 +1,80 @@
+test_covars <- function(df) {
+  d <- lag_any_candidates(DF) |>
+    dplyr::select(year, full_district, prev_dem_margin, white = prcntWhiteAll, income = meanIncome, unemp = prcntUnemp) |>
+    tidyr::pivot_longer(cols = c(white, income, unemp),
+                        names_to = "demo") |>
+    dplyr::filter(!is.na(value))
+  
+  d |>
+    dplyr::group_by(demo) |>
+    tidyr::nest(data = c(year, full_district, prev_dem_margin, value)) |>
+    dplyr::mutate(rdd = purrr::map(
+      data,
+      function(x) {
+        est <- rdd::RDestimate(value ~ prev_dem_margin, x)$est[1]
+        se <- rdd::RDestimate(value ~ prev_dem_margin, x)$se[1]
+        bw <- rdd::RDestimate(value ~ prev_dem_margin, x)$bw[1]
+        data.frame(est = est, se = se, bw = bw)
+      }
+    )) |>
+    tidyr::unnest(rdd)
+  
+  purrr::map_dfr(
+    seq(0.01, 0.15, by = 0.01),
+    function(bw) {
+      d |>
+        dplyr::filter(prev_dem_margin >= -bw,
+                      prev_dem_margin <= bw) |>
+        dplyr::group_by(demo, winner) |>
+        dplyr::summarize(value = mean(value),
+                         district_year_obs = dplyr::n()) |>
+        dplyr::ungroup() |>
+        dplyr::mutate(bandwidth = bw)
+    }
+  )
+}
+
+decade_rdd <- function(df) {
+  lag_any_candidates(df) |>
+    dplyr::mutate(decade = floor(year/10)*10) |>
+    dplyr::group_by(decade) |>
+    tidyr::nest() |>
+    dplyr::mutate(
+      res = 
+        purrr::map(
+          data,
+          function(x) {
+            mccrary <- rdd::DCdensity(x$prev_dem_margin)
+            est <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$est[1]
+            se <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$se[1]
+            bw <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$bw[1]
+            
+            data.frame(mccrary = mccrary, est = est, se = se, bw = bw)
+          }
+        )
+    ) |>
+    tidyr::unnest(res)
+}
+
[…]

### 2023-04-25 R/tables.R
clean up

@@ -44,7 +44,7 @@ decade_rdd <- function(df) {
         purrr::map(
           data,
           function(x) {
-            mccrary <- rdd::DCdensity(x$prev_dem_margin)
+            mccrary <- rdd::DCdensity(x$prev_dem_margin, plot = F)
             est <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$est[1]
             se <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$se[1]
             bw <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$bw[1]
@@ -53,7 +53,8 @@ decade_rdd <- function(df) {
           }
         )
     ) |>
-    tidyr::unnest(res)
+    tidyr::unnest(res) |>
+    dplyr::select(Decade = decade, LATE = est)
 }
 
 region_rdd <- function(df,
@@ -67,7 +68,7 @@ region_rdd <- function(df,
         purrr::map(
           data,
           function(x) {
-            mccrary <- rdd::DCdensity(x$prev_dem_margin)
+            mccrary <- rdd::DCdensity(x$prev_dem_margin, plot = F)
             est <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$est[1]
             se <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$se[1]
             bw <- rdd::RDestimate(dem_margin ~ prev_dem_margin, x)$bw[1]
@@ -76,5 +77,6 @@ region_rdd <- function(df,
           }
         )
     ) |>
-    tidyr::unnest(res)
+    tidyr::unnest(res) |>
+    dplyr::select(Division, LATE = est)
 }

### 2023-04-25 paper.qmd
done??

@@
   warning: false
+  results: false
 ---
@@
 
-lag_any_candidates(DF) |>
-  dplyr::pull(prev_dem_margin) |>
-  rdd::DCdensity()
+invisible(rdd::DCdensity(lag_any_candidates(DF)$prev_dem_margin))
 ```
@@
 With no empirical discontinuities in any of the densities under investigation discovered, we can safely assume that the continuity assumption holds in our idealized election model here and feel comfortable applying RDD to estimating effects of incumbency.
+
+### Basic RDD Results
+
+With diagnostics assessed, we ran our first RDD analyses. After applying our filters, we had 4,800 observations. We used the `rdd` package in R to conduct our analyses.
+
+We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory.
+A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$.
+This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.
+
+```{r}
+#| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.
+
+lag_any_candidates(DF) |>
+  plot_rdd()
+```
+
+We also conducted logistic regressions as part of our analyses.
+Instead of examining the Democratic margin in election $t$, we consider only the final result: a Democratic victory or loss.
+These logistic regressions support the narrative from the RDD analysis, but in a simpler setting.
+
+```{r}
+#| fig-cap: Separate logistic regressions on either side of the cutoff also show a clear discontinuity, with a 0.551 change in the probability of a Democrat winning election $t$.
+
+logistic(DF)
+```
+
+To further separate potential associative relationships from causal relationships, we reversed our analyses.
+Clearly, an election in year $t$ cannot have an effect on an election in year $t-1$ because of the temporal structure in the data.
+Our empirical results align with this theory, as there is practically no discontinuity in the RDD or logistic analyses and thus no evidence of retro-causality.
+
+```{r}
+#| fig-cap: An RDD applied in reverse — the effect of the Democratic margin in election $t$ on election $t-1$ — shows no discontinuities.
+
+backwards_rdd(DF)
+```
+
+```{r}
+#| fig-cap: A logistic regression applied in reverse also shows no discontinuities.
+
+logistic_backwards(DF)
+```
+
+### Discarded Analyses
+
+Up to this point, our analyses have examined a slightly weaker form of incumbency: party incumbency.
+That is, we looked at the Democratic margin without considering the candidates themselves.
+One Democrat may run in election $t-1$ but another in election $t$, and we would treat this identically to a situation in which the same Democrat runs in elections $t-1$ and $t$.
+We sought to narrow our analysis to eight cases:
+
+- the same Democrat runs in elections $t-1$ and $t$, regardless of the best-performing non-Democrat;
+- a different Democrat runs in elections $t-1$ and $t$, regardless of the non-Democrat;
+- the same Democrat runs in elections $t-1$ and $t$ against a different non-Democrat;
+- the same Democrat runs in elections $t-1$ and $t$ against the same non-Democrat;
+- the same non-Democrat runs in elections $t-1$ and $t$, regardless of the Democrat;
+- a different non-Democrat runs in elections $t-1$ and $t$, regardless of the Democrat;
+- the same non-Democrat runs in elections $t-1$ and $t$ against a different Democrat; and
+- the same non-Democrat runs in elections $t-1$ and $t$ against the same Democrat.
+
+However, these sub-groups fail to meet the assumptions required for RDD.
+This may be because of who runs again.
+If a Democratic candidate loses by a slim margin, they are far more likely to step away from politics than a Democrat who barely wins, even though the candidates and their districts may be statistically similar.
+We empirically do not see continuity, which has a theoretical justification.
+
+```{r}
+#| fig-cap: Density plots for the first four sub-groups show violations in the continuity assumption.
+#| fig-height: 6
+
+par(mfrow = c(2, 2))
+invisible(rdd::DCdensity(lag_same_dem(DF)$prev_dem_margin) )
+invisible(rdd::DCdensity(lag_diff_dem(DF)$prev_dem_margin))
+invisible(rdd::DCdensity(lag_same_dem_diff_rep(DF)$prev_dem_margin) )
+invisible(rdd::DCdensity(lag_same_dem_same_rep(DF)$prev_dem_margin))
+```
+
+### Additional Results
+
+We did manage to successfully divide our population into some relevant sub groups.
+First, we divided the elections by decade.
+As the table shows, the effect of a Democrat winning a close election on the Democratic margin in the next election has decreased over time.
+
+```{r}
+#| tbl-cap: The LATE has decreased over time.
+
+decade_rdd(DF) |>
+  knitr::kable()
+```
+
+Then, we considered the region of the congressional district, as defined by the U.S.
+Census Bureau.
+Many regions have similar LATE estimates.
+The Pacific region, comprising Alaska, Washington, Oregon, California, and Hawaii, has the lowest estimate.
+The West South Central, containing Oklahoma, Texas, Arkansas, and Louisiana, has the second highest LATE estimate, while New England — Maine, New Hampshire, Vermont, Massachusetts, Connecticut, and Rhode Island — has the highest at 0.269.
+
+```{r}
+#| tbl-cap: Most regions see similar LATE estimates, though New England may have a particularly strong incumbency effect.
+
+region_rdd(DF) |>
+  knitr::kable()
+```
+
+## Conclusions
+
+Our analyses support previous literature showing a strong incumbency effect.
+Districts that are otherwise similar, as our covariate analysis showed, may have vastly different electoral futures come down to just a handful of votes.
+These results show the importance of national parties investing in close congressional elections, as a marginal amount of votes in one election can affect following elections.
+
+There are many analyses researchers can conduct to further the study of incumbency advantages.
+Demographic data on the candidates themselves can be considered to explore racial and other sorts of discrimination in politics.
+Scientists can examine when a losing candidate runs again: Do they take a few cycles to recover from a grueling and heartbreaking campaign?
+Do they never re-enter politics?
+Does it depend on how close they were?
+Finally, scientists can reproduce these analyses at other levels — the U.S.
+Senate, state senates, state houses, and local offices.

### 2023-04-25 R/rdd.R
remove printing

@@ -52,10 +52,10 @@ covariate_rdd <- function(df) {
 backwards_rdd <- function(df, figs = config$paths$figures, dig = 3, cols = config$analysis$colors) {
   d <- lag_any_candidates(df)
   
-  mccrary_p <- rdd::DCdensity(d$dem_margin)
-  jpeg(paste0(figs, "backwards_rdd", "_mccrary_p_", round(mccrary_p, dig), ".jpeg"), width = 6, height = 6, units = "in", res = 300)
-  rdd::DCdensity(d$dem_margin)
-  dev.off()
+  # mccrary_p <- rdd::DCdensity(d$dem_margin)
+  # jpeg(paste0(figs, "backwards_rdd", "_mccrary_p_", round(mccrary_p, dig), ".jpeg"), width = 6, height = 6, units = "in", res = 300)
+  # rdd::DCdensity(d$dem_margin)
+  # dev.off()
   
   late <- round(rdd::RDestimate(prev_dem_margin ~ dem_margin, d)$est["LATE"], dig)
   
@@ -96,9 +96,9 @@ logistic <- function(df, cols = config$analysis$colors) {
                    family = "binomial")
   
   eps <- 0.001
-  (predict(fit_upper, newdata = data.frame(prev_dem_margin = -eps), type = "response") - 
-    predict(fit_lower, newdata = data.frame(prev_dem_margin = eps), type = "response")) |>
-    print()
+  # (predict(fit_upper, newdata = data.frame(prev_dem_margin = -eps), type = "response") - 
+  #   predict(fit_lower, newdata = data.frame(prev_dem_margin = eps), type = "response")) |>
+  #   print()
   
   d_lower |>
     dplyr::mutate(p = predict(fit_lower, newdata = d_lower, type = "response"),
@@ -140,9 +140,9 @@ logistic_backwards <- function(df, cols = config$analysis$colors) {
                    family = "binomial")
   
   eps <- 0.001
-  (predict(fit_upper, newdata = data.frame(dem_margin = -eps), type = "response") - 
-      predict(fit_lower, newdata = data.frame(dem_margin = eps), type = "response")) |>
-    print()
+  # (predict(fit_upper, newdata = data.frame(dem_margin = -eps), type = "response") - 
+  #     predict(fit_lower, newdata = data.frame(dem_margin = eps), type = "response")) |>
+  #   print()
   
   d_lower |>
     dplyr::mutate(p = predict(fit_lower, newdata = d_lower, type = "response"),

### 2023-04-25 paper.qmd
try cross ref

@@
 ---
-title: "Regression Discontinuity Design to Measure Incumbency Advantage in US Congressional Elections"
-subtitle: "Final Project for STAT 530"
+title: "Regression Discontinuity Design to Measure Incumbency Advantage in U.S. Congressional Elections"
+subtitle: "Final Project for STAT 530, Spring 2023"
 author: "Tyler Bagwell and Prayag Gordy"
-date: "April 24, 2023"
 abstract: "The incumbency effect in U.S.
@@
 format: pdf
+bibliography: 530project_bibtex.bib
 toc: true
@@
 House of Representatives are reelected at astonishing rates.
-Since 1964, the reelection rate has never dropped below 85%, while the approval rating of the House as a whole has generally remained around 40%, dipping to 20% in recent years.
-Thus, many political scientists spend their careers researching the advantages of incumbency (Gelman and King, 1990).
+Since 1964, the reelection rate has never dropped below 85% [@OS1], while the approval rating of the House as a whole has generally remained around 40%, dipping to 20% in recent years [@Gallup1].
+Thus, many political scientists spend their careers researching the advantages of incumbency [@Gelman1990].
 
 Many experts have criticized the apparently strong effects of incumbency.
-Congressmen can use government funds to visit areas of their district, meet with important political players, and send mail to constituents touting their achievements, all of which indirectly benefit their reelection campaign.
+Congressmen can use government funds to visit areas of their district, meet with important political players, and send mail to constituents touting their achievements, all of which may indirectly benefit their reelection campaign [@NYT1].
 Challengers enjoy none of these benefits.
@@
 Previous research has found strong effects of incumbency.
-Candidates running for reelection routinely raise far more money.
-Further, they may be able to deter challengers simply by being an incumbent.
-Overall, Cummings found that in similar districts, incumbent congressmen are more likely to win elections than non-incumbent congressmen from the same party.
+Candidates running for reelection routinely raise far more money [@OS2].
+Further, they may be able to deter challengers simply by being an incumbent [@Assessing2015].
+Overall, Cummings found that in similar districts, incumbent congressmen are more likely to win elections than non-incumbent congressmen from the same party [@Cummings1966].
 
@@
 In practice, it is impossible to determine if the continuity assumption holds for a particular experimental design unless the RDD practitioner has full knowledge of the conditional population densities of the covariates $\mathbf{X}$ with respect to the assigning variable $W$, i.e., $f(\mathbf{X}|W)$, for all covariates that are relevant to $Y$.
-In empirical RDD, the continuity assumption can be tested with the observed data by (1) testing for the lack of discontinuity in the density of the assignment variable $W$ at threshold $W=w_0$ (McCrary density test) and (2) testing for discontinuity in the first moment of the covariates $\mathbf{X}$ conditional on $W$ at $W=w_0$.
+In empirical RDD, the continuity assumption can be tested with the observed data by (1) testing for the lack of discontinuity in the density of the assignment variable $W$ at threshold $W=w_0$ (McCrary density test) and (2) testing for discontinuity in the first moment of the covariates $\mathbf{X}$ conditional on $W$ at $W=w_0$ [@McCrary2008].
 This second test is usually performed by using local linear regressions in the proximity of the line $W=w_0$.
@@
 The first data set, _U.S.
-House 1976–2020 Elections_, is a collection of results of US Congressional House elections between 1976 and 2020 for all congressional districts.
+House 1976–2020 Elections_, is a collection of results of US Congressional House elections between 1976 and 2020 for all congressional districts [@dataset1].
 The primary variables used in this data set were _candidate_, _candidate party_, _candidate votes_, and _total votes for election_.
-The second data set, _Historical Congressional Legislation and District Demographics 1972-2014_, contains a subset of the pretreatment covariates per congressional district we used for this study, namely, _unemployment rate_, _race_, and _level of education_.
+The second data set, _Historical Congressional Legislation and District Demographics 1972-2014_, contains a subset of the pretreatment covariates per congressional district we used for this study, namely, _unemployment rate_, _race_, and _level of education_ [@dataset2].
 
@@
 The results of the McCrary density test on the Democratic vote margin at the zero margin cutoff indicates a continuous density on both sides of the cutoff value ($\alpha = 0.05$).
-A plot of the McCrary density test’s density estimation is provided below.
+A plot of the McCrary density test’s density estimation is provided in @density-all.
 
@@
 #| fig-cap: McCrary density test and estimation for all $t-1$ Democratic vote margins.
+#| label: density-all
 
@@
 With these three pretreatment covariates chosen, we estimate the discontinuities in their conditional densities conditional on the $t-1$ Democratic election margin at the zero margin value using local linear regression and find that there are no statistically significant discontinuities in either three covariates’ densities.
-This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin.
+This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin, as in @covars.
 
@@
 #| fig-cap: Three pretreatment covariates versus $t-1$ Democratic vote margin.
+#| label: covars
 
@@
 
-We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory.
-A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$.
-This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.
-
 ```{r}
+#| label: basic-rdd
 #| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.
@@
 ```
+
+We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory.
+A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$.
+This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.
+See @basic-rdd.
 
@@
 
-backwards_rdd(DF)
+invisible(backwards_rdd(DF))
 ```
@@
 Senate, state senates, state houses, and local offices.
+
+## References

### 2023-04-25 paper.qmd
give up on cross ref

@@
 The results of the McCrary density test on the Democratic vote margin at the zero margin cutoff indicates a continuous density on both sides of the cutoff value ($\alpha = 0.05$).
-A plot of the McCrary density test’s density estimation is provided in @density-all.
+A plot of the McCrary density test’s density estimation is provided below.
 
@@
 #| fig-cap: McCrary density test and estimation for all $t-1$ Democratic vote margins.
-#| label: density-all
 
@@
 With these three pretreatment covariates chosen, we estimate the discontinuities in their conditional densities conditional on the $t-1$ Democratic election margin at the zero margin value using local linear regression and find that there are no statistically significant discontinuities in either three covariates’ densities.
-This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin, as in @covars.
+This result can also be expected just by visual inspection of the covariate value versus the $t-1$ Democratic election margin.
 
@@
 #| fig-cap: Three pretreatment covariates versus $t-1$ Democratic vote margin.
-#| label: covars
 
@@
 
+We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory.
+A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$.
+This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.
+
 ```{r}
-#| label: basic-rdd
 #| fig-cap: RDD analysis of 4,800 elections from 1976 through 2020. The local average treatment effect is 0.173.
@@
 ```
-
-We found a clear discontinuity from a slight Democratic loss to a slight Democratic victory.
-A district that votes barely against the Democrat in election $t-1$ tends to shift further away from Democrats in election $t$, while districts that only marginally elect a Democrat in election $t-1$ move toward Democrats in election $t$.
-This is a clear effect of the treatment — in this case, electing a Democrat — on the Democratic margin in the next election.
-See @basic-rdd.
 
